\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{booktabs}
\usepackage{hyperref}

\begin{document}

\begin{center}
{\LARGE\bfseries Research Reporting Guideline Statement}
\end{center}

\bigskip

\noindent\textbf{Manuscript Title:} Geometric Clinical Ethics: Medical Decision-Making as Pathfinding on the Moral Manifold, and the Mathematical Structure of Moral Injury

\noindent\textbf{Article Type:} Extended Essay

\noindent\textbf{Journal:} \emph{Journal of Medical Ethics}

\bigskip

\section*{Applicability of Standard Reporting Guidelines}

This manuscript is a \textbf{theoretical and philosophical contribution} that develops a mathematical framework for clinical ethical reasoning.  It does not report an empirical study, clinical trial, systematic review, qualitative research, case report, or any other form of primary data collection or analysis.

No human participants were recruited.  No clinical data were collected, accessed, or analysed.  No interventions were administered.  No datasets were generated.

The standard EQUATOR Network reporting guidelines are designed for empirical research and are not applicable to this work:

\bigskip

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Guideline} & \textbf{Study Type} & \textbf{Applicable?} \\
\midrule
CONSORT & Randomised controlled trials & No \\
STROBE & Observational studies & No \\
PRISMA & Systematic reviews & No \\
COREQ & Qualitative research & No \\
STARD & Diagnostic accuracy & No \\
CARE & Case reports & No \\
CHEERS & Health economic evaluations & No \\
SRQR & Qualitative research & No \\
SQUIRE & Quality improvement & No \\
\bottomrule
\end{tabular}
\end{center}

\bigskip

\section*{Transparency Statement}

Although no standard checklist applies, the manuscript adheres to the following transparency principles:

\begin{enumerate}
    \item \textbf{Reproducibility of formal results:} All theorems, propositions, and proofs are stated with full formal detail sufficient for independent verification.
    \item \textbf{Falsifiability:} The framework generates six specific, \emph{a priori} falsifiable predictions (Section~7) with explicit falsification criteria.
    \item \textbf{Open methods:} The parent mathematical framework is published and publicly available; the parent software is released under the MIT open-source licence.
    \item \textbf{Conflicts of interest:} None declared.
    \item \textbf{Funding:} None received.
    \item \textbf{AI disclosure:} Generative AI (Claude, Anthropic) was used for manuscript preparation and formatting.  All intellectual content is the author's own.
\end{enumerate}

\end{document}
