\documentclass[12pt,letterpaper]{article}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{url}
\geometry{margin=1in}

\begin{document}

\begin{titlepage}
\centering
\vspace*{2cm}

{\LARGE\bfseries Representational Consistency for Machine Learning Systems:\\
A Categorical Framework for Verifying Declared Equivalences\\
With the Bond Index as Engineering Deliverable\par}

\vspace{2cm}

{\large Andrew H. Bond, Senior Member, IEEE\par}
\vspace{0.5cm}
{\normalsize Department of Computer Engineering\\
San Jos\'{e} State University\\
\texttt{andrew.bond@sjsu.edu}\\
ORCID: 0009-0003-2599-6158\par}

\vspace{1cm}

{\normalsize Spring 2026\par}

\vfill

\noindent\rule{\textwidth}{0.4pt}
\small

\noindent\textbf{AI-Assisted Writing Disclosure:} Portions of this manuscript were drafted and refined with AI assistance (Claude, Anthropic). All technical content, mathematical results, and claims were conceived, verified, and approved by the author.

\vspace{0.5em}
\noindent\textbf{Ethics/IRB Statement:} This research involves no human subjects. Empirical validation uses publicly available corpora and synthetic data. No IRB approval was required.

\vspace{0.5em}
\noindent\textbf{Data Availability:} Code and data are available at \url{https://github.com/ahb-sjsu/erisml-lib}.

\vspace{0.5em}
\noindent\textbf{Funding/Conflicts of Interest:} The author declares no conflicts of interest. This research received no external funding.

\vspace{0.5em}
\noindent\textbf{CRediT Author Statement:} Andrew H. Bond: Conceptualization, Methodology, Formal Analysis, Investigation, Writing --- Original Draft, Writing --- Review \& Editing, Software.

\normalsize
\noindent\rule{\textwidth}{0.4pt}

\end{titlepage}

\end{document}
