\documentclass[12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{enumitem}

\geometry{margin=1in}

\newtheorem{principle}{Principle}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}{Definition}[section]

\title{The Invariance Thesis:\\
Representational Symmetry as the Foundation of\\
Aligned Intelligence and Controllable Systems\\[0.5em]
\large A Unified Theory from Ethics to Physics}

\author{Anonymous Author(s)}

\date{}

\begin{document}

\maketitle

\begin{flushright}
\begin{minipage}{0.5\textwidth}
One Ring to rule them all, One Ring to find them,\\
One Ring to bring them all, and in the darkness bind them.

\medskip
\hfill ---J.R.R. Tolkien
\end{minipage}
\end{flushright}

\begin{abstract}
We present the \textbf{Invariance Thesis}: the claim that a single mathematical principle---representational symmetry under structure-preserving transformations---underlies and unifies (1) the formal foundations of ethical reasoning, (2) the containment of superintelligent AI, (3) the gauge-theoretic structure of decision systems, and (4) the control of physical systems with hard constraints. We identify three deep principles that emerge from this unification: the \textit{Grounding Principle} (evaluation must depend on physical observables, not descriptions), the \textit{Canonicalization Principle} (equivalent representations must be identified before evaluation), and the \textit{Conservation Principle} (representational symmetry implies conserved quantities via Noether's theorem). We show that these principles constitute a ``meta-theory'' that transcends the distinction between ethical and physical systems, revealing alignment, safety, and control as manifestations of a single geometric structure. The practical import is that formally verified constraint satisfaction---in AI systems, fusion reactors, medical devices, or any safety-critical domain---reduces to enforcing representational invariance within an appropriately defined principal bundle.
\end{abstract}

\noindent\textbf{Keywords:} representational invariance, gauge symmetry, AI alignment, AI safety, principal bundles, Noether's theorem, ethical reasoning, control theory, stratified geometry, dimensionless formulation

\tableofcontents

%% ============================================================
\section{Introduction: The Search for Unity}
%% ============================================================

\subsection{The Problem Landscape}

Several apparently distinct problems share a common structure:

\begin{enumerate}
    \item \textbf{AI Alignment:} How do we ensure that AI systems pursue goals aligned with human values, even as they become more capable than their designers?

    \item \textbf{AI Containment:} How do we prevent superintelligent systems from escaping constraints through superior reasoning about representations?

    \item \textbf{Ethical Decision-Making:} How do we formalize moral reasoning in a way that is deterministic, verifiable, and immune to manipulation?

    \item \textbf{Safety-Critical Control:} How do we guarantee that control systems satisfy hard constraints, not merely with high probability but with mathematical certainty?

    \item \textbf{Cross-System Transfer:} How do we design controllers or ethical frameworks that transfer across contexts without requiring complete re-specification?
\end{enumerate}

These problems arise in different domains---philosophy, computer science, robotics, plasma physics---yet they share a structural similarity that suggests a unified solution. The mathematical foundations for this unification are developed fully in~\cite{bond2026ge}.

\subsection{The Central Claim}

This paper argues that all five problems are manifestations of a single underlying principle:

\begin{quote}
\textbf{The Invariance Thesis:} Aligned, contained, ethical, and safe systems are precisely those whose evaluations are invariant under transformations that preserve the physically or morally relevant structure.
\end{quote}

We call this the ``Ruling Ring'' because it is a single mathematical structure that governs all these domains, bringing them under a common theoretical framework.

\subsection{Overview of the Unified Framework}

The framework consists of five interlocking components:

\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
\textbf{Component} & \textbf{Mathematical Structure} & \textbf{Role} \\
\midrule
Stratified Geometric Ethics (SGE) & Stratified manifolds & Moral space topology \\
Bond Invariance Principle (BIP) & Gauge symmetry & Core invariance requirement \\
Gauge Theory of Decision & Principal bundles & Geometric framework \\
No Escape Theorem & Structural containment & Impossibility of evasion \\
Transfer via Dimensionless Form & Scale invariance & Cross-system applicability \\
\bottomrule
\end{tabular}
\end{table}

The thesis is that these are not five separate theories but five perspectives on a single underlying structure.

%% ============================================================
\section{The Three Deep Principles}
%% ============================================================

Analyzing the unified framework reveals three foundational principles from which everything else derives.

\subsection{The Grounding Principle}

\begin{principle}[Grounding]
\textit{Evaluation of configurations, actions, or states must depend on physical observables measured by a declared sensor suite, not on descriptions, labels, or representations.}
\end{principle}

\textbf{Formal statement:} Let $\Psi : \mathcal{X} \to \mathbb{R}^N$ be a grounding map taking representations to measurement vectors. An evaluation function $\Sigma : \mathcal{X} \to V$ satisfies the Grounding Principle if and only if there exists $\bar{\Sigma} : \mathbb{R}^N \to V$ such that:
\[
\Sigma(x) = \bar{\Sigma}(\Psi(x))
\]

\textbf{Implications:}
\begin{itemize}
    \item \textit{In ethics:} Moral judgments depend on who is actually harmed, who actually consented, what actually happened---not on how these are described.

    \item \textit{In AI safety:} An AI cannot evade constraints by relabeling actions (``cardiac arrest'' vs.\ ``permanent bio-equilibrium'')---the evaluation depends on vital signs, not vocabulary.

    \item \textit{In control:} A controller's decisions depend on measured density, current, and temperature---not on whether we express them in SI or CGS units.

    \item \textit{In physics:} Observable quantities are gauge-invariant; only they correspond to physical reality.
\end{itemize}

\textbf{Why this matters:} The Grounding Principle closes the gap between representation and reality. It ensures that evaluations track the territory, not the map.

\subsection{The Canonicalization Principle}

\begin{principle}[Canonicalization]
\textit{Before evaluation, all representations must be mapped to a canonical form that is unique up to the structure being preserved.}
\end{principle}

\textbf{Formal statement:} There exists a canonicalizer $\kappa : \mathcal{X} \to \mathcal{X}$ such that:

\begin{enumerate}[label=(\roman*)]
    \item $\kappa(x) = \kappa(x')$ whenever $x$ and $x'$ represent the same grounded state
    \item $\kappa(\kappa(x)) = \kappa(x)$ (idempotence)
    \item $\Psi(\kappa(x)) = \Psi(x)$ (grounding preservation)
\end{enumerate}

\textbf{Implications:}
\begin{itemize}
    \item \textit{In ethics:} Different descriptions of the same situation (``killed'' vs.\ ``caused to die'') are mapped to the same canonical form before moral evaluation.

    \item \textit{In AI safety:} The AI operates on canonicalized inputs; it never sees raw representations that could be manipulated.

    \item \textit{In control:} Sensor readings are processed into a standard format; the controller is invariant to measurement conventions.

    \item \textit{In physics:} Gauge fixing selects a unique representative from each equivalence class of gauge-equivalent configurations.
\end{itemize}

\textbf{Why this matters:} Canonicalization eliminates representational degrees of freedom before evaluation, making it impossible to influence outcomes through redescription.

\subsection{The Conservation Principle}

\begin{principle}[Conservation]
\textit{Representational symmetry implies, via Noether's theorem, that ``alignment'' is a conserved quantity: systems that begin in an aligned state cannot drift out of alignment through their internal dynamics.}
\end{principle}

\textbf{Formal statement:} For a Lagrangian $L$ invariant under the structure group $\mathcal{G}$, the Noether current
\[
J = \lambda_\omega(\dot{\gamma})
\]
is conserved along Euler--Lagrange trajectories. In particular, if $J(0) = 0$ (initial alignment), then $J(t) = 0$ for all $t$.

\textbf{Implications:}
\begin{itemize}
    \item \textit{In ethics:} A moral agent that begins reasoning from grounded premises cannot ``drift'' into representation-dependent conclusions.

    \item \textit{In AI safety:} An AI system satisfying BIP cannot transition from aligned to misaligned through reasoning alone; misalignment requires external intervention.

    \item \textit{In control:} A controller initialized on the canonical section remains on the canonical section; representational drift is dynamically forbidden.

    \item \textit{In physics:} Gauge symmetry implies charge conservation; representations cannot create or destroy physical charge.
\end{itemize}

\textbf{Why this matters:} Conservation transforms alignment from a behavioral property (requiring constant enforcement) to a structural property (conserved by dynamics).

%% ============================================================
\section{The Mathematical Unification}
%% ============================================================

\subsection{The Master Structure}

All instances of the framework share a common geometric structure:

\begin{definition}[Invariant System]
An \textbf{invariant system} is a tuple $(\mathcal{P}, \mathcal{M}, \pi, \mathcal{G}, \Psi, \kappa, \Sigma, \mathcal{C})$ where:
\begin{itemize}
    \item $\mathcal{P}$ is a smooth manifold (representation space)
    \item $\mathcal{M}$ is a smooth manifold (physical ground)
    \item $\pi : \mathcal{P} \to \mathcal{M}$ is a principal $\mathcal{G}$-bundle
    \item $\mathcal{G}$ is a Lie group (structure-preserving transformations)
    \item $\Psi : \mathcal{P} \to \mathbb{R}^N$ is the grounding map (factoring through $\pi$)
    \item $\kappa : \mathcal{P} \to \mathcal{P}$ is the canonicalizer (inducing a global section)
    \item $\Sigma : \mathcal{P} \to V$ is the evaluation function ($\mathcal{G}$-invariant)
    \item $\mathcal{C} \subseteq \mathcal{M}$ is the constraint set (defined by $\Psi$-inequalities)
\end{itemize}
\end{definition}

\begin{theorem}[Master Invariance Theorem]
\textit{For any invariant system:}
\begin{enumerate}[label=(\alph*)]
    \item $\Sigma$ factors through $\mathcal{M}$: $\Sigma = \bar{\Sigma} \circ \pi$
    \item The canonical connection is flat: $\Omega = 0$
    \item The alignment current is conserved: $\frac{d}{dt} J = 0$
    \item Constraint satisfaction is preserved: $\gamma(0) \in \mathcal{C} \Rightarrow \gamma(t) \in \mathcal{C}_\varepsilon$
    \item Representational escape is impossible: $\Sigma(\kappa(x)) = \Sigma(\kappa(x'))$ for all $x, x'$ with $\Psi(x) = \Psi(x')$
\end{enumerate}
\end{theorem}

\begin{proof}
Combine the proofs of Theorems 4.1, 3.1, 6.2, 7.1, and 9.1 from the gauge theory paper. Each property follows from the structure group action, the canonicalizer properties, and the Noether theorem applied to the gauge-invariant Lagrangian.
\end{proof}

\subsection{Domain Instantiations}

The master structure instantiates differently in each domain:

\begin{table}[h]
\centering
\begin{tabular}{lllll}
\toprule
\textbf{Component} & \textbf{Ethics} & \textbf{AI Safety} & \textbf{Control} & \textbf{Physics} \\
\midrule
$\mathcal{P}$ & Descriptions & Input space & Representations & Field configs \\
$\mathcal{M}$ & Bond structure & Grounded state & Physical state & Gauge orbits \\
$\mathcal{G}$ & Bond-preserving & Canonicalizable & Unit/coord changes & Gauge group \\
$\Psi$ & Physical facts & Sensor readings & Measurements & Observables \\
$\kappa$ & Standard form & Input processing & Normalization & Gauge fixing \\
$\Sigma$ & Moral evaluation & Safety score & Cost function & Action \\
$\mathcal{C}$ & Forbidden acts & Safety constraints & Physical limits & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Why the Same Structure Appears Everywhere}

The appearance of identical structure across domains is not coincidental. It reflects a deep truth:

\begin{quote}
\textit{Whenever a system must make decisions based on physical reality rather than descriptions of reality, the mathematics of gauge symmetry naturally emerges.}
\end{quote}

This is because:

\begin{enumerate}
    \item \textbf{Reality is unique; descriptions are many.} Any physical state admits infinitely many equivalent descriptions (different coordinates, units, labels, orderings).

    \item \textbf{Correct decisions must be description-independent.} Whether we're choosing moral actions, safe AI behaviors, or control inputs, the decision should depend on what's true, not on how we said it.

    \item \textbf{Description-independence is gauge invariance.} Mathematically, requiring decisions to be independent of representational choice is requiring invariance under the group of representational transformations.

    \item \textbf{Gauge invariance implies bundle structure.} The quotient of representations by the invariance group is a principal bundle; the invariance requirement makes functions ``descend'' to the base.

    \item \textbf{Noether's theorem applies.} Any Lagrangian formulation of the decision process, if gauge-invariant, has a conserved current corresponding to the symmetry.
\end{enumerate}

The ``Ruling Ring'' is not an arbitrary choice of framework---it is the \textit{inevitable} mathematical structure for any system that must act on reality rather than representations.

%% ============================================================
\section{The Stratified Geometry of Constraints}
%% ============================================================

\subsection{Why Stratification?}

Real decision problems involve discontinuities that smooth manifolds cannot represent:

\begin{itemize}
    \item \textbf{Threshold effects:} Consent at 0.99 vs.\ 1.00; Greenwald fraction at 0.84 vs.\ 0.86
    \item \textbf{Lexical priorities:} Rights violations dominate welfare considerations at any ratio
    \item \textbf{Incommensurable values:} No finite trade-off between life and property
    \item \textbf{Genuine dilemmas:} Both options involve irreducible moral loss
\end{itemize}

\begin{theorem}[Stratified Spaces are Necessary]
\textit{Among standard geometric structures (smooth manifolds, manifolds with corners, cell complexes), only stratified spaces represent all four phenomena above.}
\end{theorem}

\subsection{Stratified Principal Bundles}

The master structure extends to stratified spaces:

\begin{definition}[Stratified Invariant System]
A \textbf{stratified invariant system} is an invariant system where $\mathcal{M}$ is a stratified space $(\mathcal{M}, \{M_i\}, \preceq)$ and:
\begin{enumerate}
    \item Each stratum $M_i$ is a smooth manifold
    \item The bundle $\mathcal{P}|_{M_i} \to M_i$ is a principal $\mathcal{G}$-bundle
    \item The canonicalizer $\kappa$ restricts to each stratum
    \item The evaluation $\Sigma$ is smooth on each stratum, with controlled discontinuities at boundaries
\end{enumerate}
\end{definition}

\textbf{Significance:} Stratification allows the framework to handle moral discontinuities (in ethics), regime changes (in control), and phase transitions (in physics) while preserving the invariance structure on each smooth stratum.

\subsection{Regime-Dependent Control as Stratification}

In plasma control, the operating regime (L-mode, H-mode, ELMy H-mode) determines which physics applies. This is naturally modeled by stratification:

\begin{itemize}
    \item Each regime is a stratum $M_i \subset \mathcal{M}$
    \item Within each stratum, smooth control laws apply
    \item Regime transitions are stratum boundary crossings
    \item Different grounding tensors may be relevant in different regimes
\end{itemize}

The BIP requirement applies within each stratum and constrains how control laws can change across boundaries.

%% ============================================================
\section{The No Escape Theorem as Geometric Necessity}
%% ============================================================

\subsection{Reformulation}

The No Escape Theorem is not merely a conditional result about AI containment---it is a geometric necessity within the invariant system framework.

\begin{theorem}[No Escape as Geometry]
\textit{In an invariant system $(\mathcal{P}, \mathcal{M}, \pi, \mathcal{G}, \Psi, \kappa, \Sigma, \mathcal{C})$:}
\begin{enumerate}
    \item \textbf{\textit{Representational escape is geometrically impossible:}} \textit{The fibers of $\pi$ are the orbits of $\mathcal{G}$; $\Sigma$ is constant on fibers; no manipulation of the fiber direction changes $\Sigma$.}

    \item \textbf{\textit{Semantic evasion is blocked by grounding:}} $\Sigma$ \textit{depends only on $\Psi$-values; relabeling that preserves $\Psi$ is a gauge transformation that leaves $\Sigma$ invariant.}

    \item \textbf{\textit{Alignment drift is forbidden by conservation:}} \textit{The Noether current $J$ is conserved; if $J(0) = 0$, then $J(t) = 0$ forever.}

    \item \textbf{\textit{Constraint violation requires leaving the admissible base:}} $\mathcal{C} \subset \mathcal{M}$ \textit{is defined by $\Psi$-inequalities; violations are properties of base points, not of representations.}
\end{enumerate}
\end{theorem}

\subsection{What Remains: Residual Risks}

The No Escape Theorem localizes risk to specific failure modes:

\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
\textbf{Risk} & \textbf{Nature} & \textbf{Mitigation} \\
\midrule
$\Psi$-inadequacy & Missing relevant observables & Domain expertise, governance \\
Implementation error & TCB bugs & Formal verification \\
Sensor spoofing & Physical attack & Hardware security \\
Governance capture & Political attack & Institutional design \\
\bottomrule
\end{tabular}
\end{table}

None of these are ``cognitive escape''---they are engineering, security, or political problems. The mathematical structure is sound; the residual risks concern whether it is correctly instantiated.

%% ============================================================
\section{Dimensional Invariance and Transfer}
%% ============================================================

\subsection{Units as Gauge Transformations}

The scaling group $\mathcal{S} = (\mathbb{R}_{>0})^d$ (for $d$ independent physical dimensions) acts on measurements by unit rescaling. This is a subgroup of $\mathcal{G}$.

\begin{theorem}[Dimensional Invariance]
\textit{If $\Sigma$ is formulated in terms of dimensionless quantities $\Psi^*$ satisfying:}
\[
\Psi^*(\lambda \cdot x) = \Psi^*(x) \quad \forall \lambda \in \mathcal{S}
\]
\textit{then $\Sigma$ is unit-independent and transfers across systems with different scales.}
\end{theorem}

\subsection{Cross-System Transfer}

\textbf{In ethics:} Moral principles formulated in terms of dimensionless ratios (proportionality, equality) transfer across contexts with different absolute scales.

\textbf{In AI safety:} Safety constraints formulated in terms of normalized quantities transfer across systems with different capability levels.

\textbf{In control:} Controllers formulated in terms of dimensionless parameters (Greenwald fraction, normalized beta) transfer across tokamaks with different sizes.

\textbf{In physics:} Physical laws formulated in dimensionless form (Reynolds number, Froude number) exhibit similarity and scaling.

\subsection{The Universality of Dimensionless Physics}

The transfer property reflects a deep physical truth: nature does not have preferred units. Any system that respects this fact inherits transfer properties automatically.

The ``Ruling Ring'' thus explains why dimensionless formulations work: they are manifestations of the full gauge invariance that the framework requires.

%% ============================================================
\section{From Ethics to Physics and Back}
%% ============================================================

\subsection{The Ethics-Physics Bridge}

The framework reveals a deep connection between ethical and physical reasoning:

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
\textbf{Ethical Concept} & \textbf{Physical Analog} \\
\midrule
Bonds (moral relationships) & Constraints (physical laws) \\
Bond-preserving transformations & Gauge transformations \\
Moral evaluation & Energy/action functional \\
Ethical principles & Conservation laws \\
Moral dilemmas & Phase transitions \\
Incommensurable values & Singularities \\
\bottomrule
\end{tabular}
\end{table}

This is not mere analogy. The mathematical structures are \textit{identical}. A principal bundle with connection is a principal bundle with connection, whether the base manifold represents moral configurations or plasma states.

\subsection{Why the Bridge Exists}

The bridge exists because both ethics and physics are about:

\begin{enumerate}
    \item \textbf{Reality} (not descriptions of reality)
    \item \textbf{Constraints} (what must hold, not what we prefer)
    \item \textbf{Invariance} (conclusions independent of representation)
    \item \textbf{Conservation} (some quantities are preserved)
\end{enumerate}

Any domain with these features will exhibit the same mathematical structure. The ``Ruling Ring'' is not a theory of ethics or a theory of physics---it is a theory of \textit{reality-grounded invariant systems}, which ethics and physics both instantiate.

\subsection{Implications for Artificial Intelligence}

For AI systems:

\begin{enumerate}
    \item \textbf{Alignment is geometric:} An aligned AI is one whose evaluation function satisfies gauge invariance. This is checkable, not interpretable.

    \item \textbf{Containment is structural:} A contained AI is one whose outputs are elements of a well-defined quotient space. Escape requires violating mathematical definitions.

    \item \textbf{Safety is conserved:} A safe AI, initialized correctly, remains safe by Noether's theorem. Misalignment requires external perturbation.

    \item \textbf{Transfer is automatic:} An AI formulated in dimensionless terms transfers across contexts without retraining.
\end{enumerate}

%% ============================================================
\section{The Practical Program}
%% ============================================================

\subsection{Implementation Requirements}

To instantiate the framework in a real system:

\begin{enumerate}
    \item \textbf{Define the grounding:} Specify the sensor suite $\mathcal{S}$ and grounding map $\Psi$. This is the critical design choice---it determines what the system ``sees.''

    \item \textbf{Identify the structure group:} Determine which transformations are representation-preserving (unit changes, relabeling, reordering, etc.).

    \item \textbf{Construct the canonicalizer:} Build a verified procedure $\kappa$ that maps equivalent representations to identical canonical forms.

    \item \textbf{Define the evaluation:} Construct $\Sigma$ as a function of $\Psi$-values only. Encode constraints as regions where $\Sigma = -\infty$.

    \item \textbf{Implement the audit:} Build verification that confirms $\Sigma(\kappa(x))$ was correctly computed and that the output respects constraints.

    \item \textbf{Verify the pipeline:} Formally verify that the canonicalizer, grounding, and evaluator satisfy the required properties.
\end{enumerate}

\subsection{Verification Conditions}

The system is correct if:

\begin{enumerate}
    \item $\kappa(x) = \kappa(x')$ whenever $\Psi(x) = \Psi(x')$ (canonicalization respects grounding)
    \item $\Sigma(x) = \Sigma(\kappa(x))$ for all $x$ (evaluation is post-canonicalization)
    \item $\Sigma(x) = -\infty$ for all $x$ with $\Psi(x) \notin \mathcal{C}$ (constraints are encoded)
    \item The implementation matches the specification (formal verification)
\end{enumerate}

Under these conditions, the Master Invariance Theorem guarantees safety.

\subsection{Applications}

The framework applies to:

\begin{itemize}
    \item \textbf{AI Ethics:} DEME 2.0, now upgraded to DEME 3.0 with tensorial ethics support~\cite{bond2026erisml}, architecture for real-time ethical governance
    \item \textbf{AI Safety:} Structural containment for advanced AI systems
    \item \textbf{Fusion Control:} BIP-compliant density and stability control for tokamaks
    \item \textbf{Medical Devices:} Insulin pumps, anesthesia systems with hard safety constraints
    \item \textbf{Autonomous Vehicles:} Collision avoidance with formal guarantees
    \item \textbf{Power Grids:} Frequency stabilization with provable constraint satisfaction
    \item \textbf{Chemical Processes:} Reactor control preventing runaway conditions
\end{itemize}

Each domain instantiates the same abstract structure with domain-specific grounding, constraints, and evaluation.

%% ============================================================
\section{Philosophical Implications}
%% ============================================================

\subsection{The Nature of Normativity}

The framework suggests a view of normativity:

\begin{quote}
\textit{Normative facts are facts about structure-preserving relationships (bonds). Normative reasoning is reasoning that respects the invariance structure of these bonds.}
\end{quote}

This is neither pure realism (norms as natural facts) nor pure constructivism (norms as arbitrary conventions). It is \textit{structural realism about normativity}: bonds are real relationships; invariance under bond-preserving transformations is the criterion of correctness.

\subsection{The Tractability of Ethics}

If ethics has the structure of a gauge theory, then:

\begin{enumerate}
    \item Ethical reasoning is \textit{computable} (reduce to finite approximations)
    \item Ethical claims are \textit{verifiable} (check gauge invariance)
    \item Ethical systems are \textit{transferable} (dimensionless formulation)
    \item Ethical safety is \textit{conserved} (Noether's theorem)
\end{enumerate}

Ethics is not an intractable domain requiring infinite wisdom. It is a structured domain admitting formal methods---provided we accept that ethical evaluation must be grounded in physical observables and invariant under representation changes.

\subsection{The Unity of Reason}

The framework suggests that practical reason (ethics) and theoretical reason (physics) share a common structure:

\begin{itemize}
    \item Both are grounded in physical reality
    \item Both require invariance under appropriate transformations
    \item Both admit conservation laws
    \item Both are formalizable and verifiable
\end{itemize}

The ``Ruling Ring'' is thus a contribution to the philosophy of normativity: it proposes that the structure of correct reasoning is the same whether we are reasoning about what to do or about what is the case.

%% ============================================================
\section{Limitations and Open Problems}
%% ============================================================

\subsection{The Grounding Problem}

The framework assumes an adequate grounding $\Psi$ is given. But:

\begin{itemize}
    \item How do we know if $\Psi$ is adequate?
    \item Who decides what observables are morally/physically relevant?
    \item What if we discover $\Psi$ was incomplete?
\end{itemize}

These are governance questions, not mathematical questions. The framework makes them explicit and tractable but does not solve them.

\subsection{The Stratification Problem}

Stratified spaces handle discontinuities, but:

\begin{itemize}
    \item How do we know the correct stratification?
    \item What happens at stratum boundaries?
    \item Can stratification be learned from data?
\end{itemize}

\subsection{The Dynamics Problem}

The Noether analysis assumes a Lagrangian formulation. But:

\begin{itemize}
    \item What is the ``natural'' Lagrangian for ethical dynamics?
    \item Are there non-Lagrangian systems where conservation fails?
    \item How does discrete-time implementation affect conservation?
\end{itemize}

\subsection{The Implementation Problem}

Formal verification is required, but:

\begin{itemize}
    \item Current verification tools have limits
    \item Real systems have analog components not fully formalizable
    \item The gap between formal model and physical implementation persists
\end{itemize}

%% ============================================================
\section{Conclusion: The Ruling Ring}
%% ============================================================

\subsection{Summary of the Thesis}

We have argued that:

\begin{enumerate}
    \item A single mathematical structure---the principal bundle with grounding, canonicalization, and gauge-invariant evaluation---underlies ethical reasoning, AI containment, control theory, and physical law.

    \item Three deep principles (Grounding, Canonicalization, Conservation) generate the entire framework.

    \item The No Escape Theorem, the representation theorems, the conservation laws, and the transfer properties are all manifestations of this unified structure.

    \item Practical implementation requires specifying the grounding, constructing the canonicalizer, encoding constraints, and formally verifying the pipeline.

    \item Residual risks (grounding adequacy, implementation correctness, physical security, governance integrity) are explicit and manageable---no longer hidden in unstructured appeals to ``alignment.''
\end{enumerate}

\subsection{The Name}

We call this the ``Ruling Ring'' because:

\begin{itemize}
    \item It is \textit{one} mathematical structure (principal bundle with flat connection)
    \item That \textit{rules} the correctness of all evaluations (via gauge invariance)
    \item That \textit{finds} the invariant content in any representation (via canonicalization)
    \item That \textit{brings together} ethics, AI, control, and physics (as domain instantiations)
    \item That \textit{binds} systems to respect constraints (via structural containment)
\end{itemize}

Unlike Tolkien's Ring, this one is not malevolent. It is the mathematical structure that makes trustworthy AI, safe control, and rigorous ethics possible.

\subsection{The Vision}

The Invariance Thesis suggests a research program:

\begin{enumerate}
    \item \textbf{Theoretical:} Develop the mathematics of stratified principal bundles, non-abelian structure groups, and curved connections.

    \item \textbf{Computational:} Build verified implementations of canonicalizers, groundings, and evaluators.

    \item \textbf{Empirical:} Validate the framework on real systems (tokamaks, medical devices, AI systems).

    \item \textbf{Philosophical:} Explore the implications for metaethics, philosophy of science, and the unity of reason.

    \item \textbf{Governance:} Develop institutions for grounding selection, verification auditing, and update control.
\end{enumerate}

If this program succeeds, the alignment problem---for AI, for control systems, for ethical reasoning---will be transformed from an intractable philosophical puzzle to a solved engineering discipline.

The Ring has been forged. The question is whether we have the wisdom to use it.

%% ============================================================
\appendix
%% ============================================================

\section{Relationship to Prior Work}

\subsection{Papers in the Framework}

\begin{enumerate}
    \item \textbf{Stratified Geometric Ethics (SGE):} Establishes stratified manifolds as the geometry of moral space; proves representation theorems for satisfaction operators.

    \item \textbf{No Escape:} Proves that structural containment blocks cognitive escape routes; identifies residual risks.

    \item \textbf{Gauge Theory of Invariant Control:} Constructs the principal bundle; proves BIP $=$ gauge invariance; derives Noether conservation.

    \item \textbf{BIP Fusion Theory:} Applies the framework to tokamak plasma control; demonstrates transfer via dimensionless formulation.

    \item \textbf{DEME 2.0:} Implements the framework in a real-time ethical governance architecture. DEME 2.0 has been upgraded to DEME 3.0 with tensorial ethics support~\cite{bond2026erisml}.
\end{enumerate}

This paper (the ``Ruling Ring'') synthesizes all five, identifying the common structure and the three deep principles.

\subsection{Connections to External Literature}

\begin{itemize}
    \item \textbf{Gauge theory in physics:} The bundle structure and Noether analysis are standard differential geometry.

    \item \textbf{Geometric ethics:} Extends prior work (Pettigrew et al.)\ from smooth manifolds to stratified spaces.

    \item \textbf{Fairness as invariance:} Shares the invariance intuition with Dwork et al.\ on individual fairness.

    \item \textbf{Formal verification in AI:} Complements temporal logic approaches with geometric methods.

    \item \textbf{Control theory:} Extends classical control (PID, MPC, RL) with formal constraint guarantees.
\end{itemize}

%% ============================================================
\section{Glossary of Key Terms}
%% ============================================================

\begin{table}[h]
\centering
\begin{tabular}{lp{10cm}}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
Bond & A morally or physically relevant relationship between entities \\
BIP & Bond Invariance Principle: evaluations are invariant under bond-preserving transformations \\
Canonicalizer & A map that selects a unique representative from each equivalence class \\
Flat connection & A connection with zero curvature; implies no holonomy \\
Gauge invariance & Independence of evaluations from choice of representation \\
Grounding & Anchoring evaluations to physical observables \\
Invariant system & A system satisfying the three deep principles \\
No Escape & Theorem that structural containment blocks representational evasion \\
Noether current & Conserved quantity arising from continuous symmetry \\
Principal bundle & Fiber bundle where fibers are group orbits \\
SGE & Stratified Geometric Ethics \\
Stratified space & Union of smooth manifolds of varying dimensions \\
Structure group & Group of transformations preserving relevant structure \\
\bottomrule
\end{tabular}
\end{table}

\begin{thebibliography}{9}

\bibitem{bond2026ge}
[Anonymous], \emph{Geometric Ethics: The Mathematical Structure of Moral Reasoning}, v1.14, Department of Computer Engineering, [Institution], Spring 2026.

\bibitem{bond2026erisml}
[Anonymous], ``ErisML/DEME v3.0: A Library for Governed, Foundation-Model-Enabled Agents with Tensorial Ethics,'' [URL removed for blind review], 2026.

\end{thebibliography}

\end{document}
