\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}

\begin{document}

\begin{center}
{\LARGE\bfseries Title Page}
\end{center}

\bigskip

\noindent\textbf{Title:} Algorithmic Jurisprudence: The Topology of Law, Constitutionality as Homology, and Legal Reasoning as Optimal Pathfinding

\bigskip

\noindent\textbf{Author:}\\
Andrew H. Bond\\
Department of Computer Engineering\\
San Jos\'{e} State University\\
San Jos\'{e}, CA 95192, USA\\
Email: \href{mailto:andrew.bond@sjsu.edu}{andrew.bond@sjsu.edu}\\
ORCID: \href{https://orcid.org/0009-0003-2599-6158}{0009-0003-2599-6158}

\bigskip

\noindent\textbf{Acknowledgments:}\\
The author thanks Lucas Thiele (UCLA) for conducting the independent replication study whose results are cited in this work.

\bigskip

\section*{Statements and Declarations}

\subsection*{Competing Interests}
The author has no relevant financial or non-financial interests to disclose.

\subsection*{Funding}
No funding was received for conducting this study.

\subsection*{Data Availability}
The reference implementation, empirical datasets, and all experimental scripts from the parent framework are publicly available at a URL withheld for double-blind review and via a public package repository. The cross-lingual corpus data and scoring pipelines are included in the repository. The legal calibration pipeline described in Section~8 has not yet been executed; the datasets required (CourtListener, Caselaw Access Project) are publicly available.

\subsection*{Code Availability}
The parent framework's library is open-source under the MIT license. The legal extension module described in Section~10 is planned but not yet implemented. Repository details are withheld for double-blind review.

\subsection*{Author Contributions}
A.H.B. conceived the framework, developed the mathematical theory, designed the empirical calibration pipeline, and wrote the manuscript.

\subsection*{Use of AI Tools}
AI-assisted copy editing tools were used to improve readability and style during manuscript preparation. All theoretical content, mathematical proofs, framework design, and scientific conclusions are the sole responsibility of the author.

\end{document}
