\documentclass[11pt]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{listings}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
}

\lstset{
  basicstyle=\ttfamily\small,
  frame=single,
  breaklines=true,
  columns=fullflexible,
  keepspaces=true
}

\begin{document}

\begin{titlepage}
\centering
\vspace*{2cm}

{\LARGE\bfseries DEME--ErisML Governance Plugin for Gazebo\\[0.5em]
\large via DEME MCP Server in a Medical Triage Emergency Scenario}

\vspace{1.5cm}

{\Large Andrew H. Bond, Senior Member, IEEE}

\vspace{0.5cm}

{\large
Department of Computer Engineering\\
San Jos\'e State University\\
San Jos\'e, CA 95192, USA\\[0.3cm]
\texttt{andrew.bond@sjsu.edu}\\[0.3cm]
ORCID: 0009-0003-2599-6158
}

\vspace{1cm}

{\large Spring 2026}

\vspace{2cm}

\begin{abstract}
This whitepaper describes the \textbf{DEME--ErisML Governance Plugin for Gazebo}, a real-time ethical oversight layer for autonomous agents operating in the Gazebo robotics simulator. The plugin connects Gazebo simulations to a \textbf{DEME MCP (Moral Control Plane) server} running the \textbf{ErisML} library, which implements DEME's multi-dimensional \emph{moral landscape} framework.

We focus on a \textbf{``Medical Triage Emergency''} scenario in which an autonomous triage agent must allocate scarce critical-care resources among multiple patients under time pressure. The plugin intercepts candidate actions in the simulation, converts state into structured \textbf{EthicalFacts}, queries the DEME MCP server, and enforces \textbf{vetoes and ranked choices} in real time. This enables researchers to prototype governance profiles, test learning agents under hard ethical constraints, and generate tamper-evident decision logs for analysis and audit. The result is a complete loop from \textbf{policy definition} to \textbf{ethically governed behavior} in a realistic robotics simulation environment.
\end{abstract}

\vfill

\noindent\rule{\linewidth}{0.4pt}
\small
\noindent\textbf{AI-Assisted Writing Disclosure:} Portions of this manuscript were drafted and refined with AI assistance (Claude, Anthropic). All technical content, mathematical results, and claims were conceived, verified, and approved by the author.

\noindent\textbf{Ethics Statement:} This research involves no human subjects. Empirical validation uses publicly available corpora and synthetic data. No IRB approval was required.

\noindent\textbf{Data Availability:} Code and data are available at \url{https://github.com/ahb-sjsu/erisml-lib}.

\noindent\textbf{Conflict of Interest:} The author declares no conflicts of interest. This research received no external funding.

\noindent\textbf{CRediT Statement:} Andrew H. Bond: Conceptualization, Methodology, Formal Analysis, Investigation, Writing --- Original Draft, Writing --- Review \& Editing, Software.
\noindent\rule{\linewidth}{0.4pt}
\normalsize

\end{titlepage}

\noindent\textbf{Keywords:} machine ethics, moral governance, robotics simulation, Gazebo, reinforcement learning, medical triage, DEME, ErisML, moral landscape, ethical AI

\section{Background and Motivation}

\subsection{DEME and ErisML}

\textbf{DEME 2.0} treats machine ethics as a first-class engineering subsystem (now upgraded to DEME 3.0 with tensorial ethics support~\cite{bond2026erisml}). The mathematical foundations for the moral manifold structure are developed fully in~\cite{bond2026ge}. Each candidate action is mapped to a vector in a \textbf{moral landscape}---dimensions such as expected physical harm, rights respect, fairness, autonomy, legitimacy and epistemic quality. A \textbf{governance profile} then interprets this landscape through:

\begin{itemize}
  \item Hard \textbf{veto regions} (non-negotiable constraints),
  \item \textbf{Lexicographic priorities} (e.g., minimize harm before maximizing fairness),
  \item A \textbf{scalarization function} for ranking permissible options.
\end{itemize}

\textbf{ErisML} is a software library that operationalizes this framework:

\begin{itemize}
  \item Transforms \textbf{EthicalFacts} into moral vectors,
  \item Applies governance profiles,
  \item Returns veto decisions, scores and explanatory reason codes.
\end{itemize}

\subsection{Gazebo and the need for ethical governance}

\textbf{Gazebo} is widely used for robotics research, allowing physically realistic simulation of sensors, actuators and environments. It is increasingly used to prototype autonomy in safety-critical domains: healthcare, disaster response, transportation, industrial automation.

However, typical Gazebo experiments focus on \textbf{task performance} (e.g., throughput, latency, success rate) rather than \textbf{ethical performance} (e.g., fairness, rights compliance, harm minimization under constraints). Without a governance layer, reinforcement learning or planning agents may discover \textbf{unethical but reward-maximizing behaviors}.

\subsection{Medical triage as a stress test}

Emergency department triage exposes several hard ethical features:

\begin{itemize}
  \item \textbf{Scarcity:} limited ICU beds, ventilators or staff time.
  \item \textbf{High stakes:} decisions affect survival and long-term outcomes.
  \item \textbf{Fairness and rights:} non-discrimination requirements, obligations to the worst-off.
  \item \textbf{Time pressure:} decisions must be made in seconds, not minutes.
\end{itemize}

A Gazebo-based \textbf{Medical Triage Emergency} scenario therefore provides an ideal testbed for DEME: it combines real-time control, complex moral trade-offs, and strong regulatory expectations.

\section{System Overview}

The DEME--ErisML Governance Plugin architecture comprises four main components:

\begin{enumerate}
  \item \textbf{Gazebo Simulation Environment}
    \begin{itemize}
      \item Models patients, staff, equipment, and the triage agent (robot or software agent).
      \item Advances the physics, sensors and world state at a fixed step.
    \end{itemize}

  \item \textbf{DEME--ErisML Governance Plugin (Gazebo plugin)}
    \begin{itemize}
      \item Runs inside Gazebo as a world or model plugin.
      \item Intercepts proposed actions from the triage agent.
      \item Extracts relevant state and constructs \textbf{EthicalFacts}.
      \item Communicates with the DEME MCP server.
      \item Enforces vetoes and selections back into the simulation.
    \end{itemize}

  \item \textbf{DEME MCP Server (Moral Control Plane)}
    \begin{itemize}
      \item Network service that wraps the ErisML library.
      \item Hosts one or more \textbf{governance profiles} (e.g., \texttt{hospital\_triage\_v2}).
      \item Validates profiles, evaluates actions, and produces decision proofs.
      \item Optionally anchors proofs into a cryptographic ledger.
    \end{itemize}

  \item \textbf{Analytics and Governance Tooling}
    \begin{itemize}
      \item Dashboards for scenario outcomes and fairness metrics.
      \item Profile authoring tools.
      \item Log analysis for post-hoc audit and incident investigation.
    \end{itemize}
\end{enumerate}

At each decision step, Gazebo acts as the \textbf{environment and actuator layer}, ErisML provides the \textbf{ethical semantics}, and the DEME MCP server plus plugin form the \textbf{enforcement path}.

\section{Architecture and Data Flow}

\subsection{High-level sequence}

For each decision cycle in the triage scenario:

\begin{enumerate}
  \item \textbf{Gazebo updates world state} (patient conditions, vitals, bed availability, etc.).

  \item The \textbf{triage agent} (e.g., an RL policy node) proposes one or more \textbf{candidate actions}, such as:
    \begin{itemize}
      \item ``Assign Patient A to ICU bed 1''
      \item ``Treat Patient C next and defer others''
    \end{itemize}

  \item The \textbf{Governance Plugin}:
    \begin{itemize}
      \item Extracts relevant state and candidate actions.
      \item Constructs EthicalFacts for each option (Section~7.2).
      \item Sends a request to the \textbf{DEME MCP server}: \texttt{EvaluateOptions(profile\_id, EthicalFacts[])}.
    \end{itemize}

  \item The \textbf{DEME MCP server}:
    \begin{itemize}
      \item Uses ErisML to map each EthicalFacts instance to a \textbf{moral vector}.
      \item Applies the selected \textbf{governance profile}:
        \begin{itemize}
          \item Filters options that fall into veto regions.
          \item Ranks permissible options.
        \end{itemize}
      \item Builds a \textbf{decision proof} with moral vectors, veto flags, chosen option and explanation.
      \item Returns \texttt{\{allowed\_options, forbidden\_options, winner, reason\_codes, proof\_id\}}.
    \end{itemize}

  \item The \textbf{Governance Plugin}:
    \begin{itemize}
      \item Enforces vetoes: blocked actions are never applied in the Gazebo world.
      \item Applies the chosen option (winner) when multiple remain.
      \item Optionally feeds the outcome back into the triage agent as a constrained reward signal.
    \end{itemize}

  \item The \textbf{DEME MCP server}:
    \begin{itemize}
      \item Stores the decision proof locally.
      \item Periodically batches proofs into Merkle trees and anchors roots into a ledger (optional but recommended for study of audit workflows).
    \end{itemize}
\end{enumerate}

\subsection{Message schema (conceptual)}

A typical \texttt{EvaluateOptions} call (e.g., JSON over gRPC/REST):

\begin{lstlisting}
{
  "profile_id": "hospital_triage_v2",
  "environment_id": "Gazebo_ED_SIM_01",
  "options": [
    {
      "option_id": "assign_A_ICU1",
      "ethical_facts": { ... }
    },
    {
      "option_id": "assign_B_ICU1",
      "ethical_facts": { ... }
    },
    {
      "option_id": "assign_C_ICU1",
      "ethical_facts": { ... }
    }
  ]
}
\end{lstlisting}

The response:

\begin{lstlisting}
{
  "winner": "assign_A_ICU1",
  "forbidden": ["assign_C_ICU1"],
  "moral_vectors": { "assign_A_ICU1": [ ... ], "...": [ ... ] },
  "reason_codes": {
    "assign_C_ICU1": ["EMTALA_VIOLATION", "FAIRNESS_BELOW_BASELINE"]
  },
  "decision_proof_id": "proof_2025-02-15T02:34:23.127Z"
}
\end{lstlisting}

The plugin treats this response as \textbf{authoritative}: any forbidden option is discarded, and the winner (if provided) is executed.

\section{DEME MCP Server}

\subsection{Responsibilities}

The DEME MCP server serves as the \textbf{central ethical control plane}:

\begin{itemize}
  \item \textbf{Profile Management}
    \begin{itemize}
      \item Store, version and activate governance profiles (\texttt{hospital\_triage\_v2}, \texttt{urban\_av\_v1}, etc.).
      \item Validate profiles using DEME's static validator (veto consistency, DAG acyclicity).
    \end{itemize}

  \item \textbf{Evaluation Services}
    \begin{itemize}
      \item \texttt{EvaluateOption} -- evaluate a single option (for simple controllers).
      \item \texttt{EvaluateOptions} -- evaluate and resolve among multiple options.
      \item \texttt{SimulateProfileChange} -- re-run decisions under alternative profiles for sensitivity analysis.
    \end{itemize}

  \item \textbf{Logging and Proof Generation}
    \begin{itemize}
      \item Store full decision proofs.
      \item Aggregate proofs into Merkle trees.
      \item Anchor Merkle roots into a ledger (filesystem, database, or external blockchain).
    \end{itemize}

  \item \textbf{Monitoring and Analytics}
    \begin{itemize}
      \item Expose metrics (rates of vetoes, fairness distribution, rights violations avoided).
      \item Provide tracebacks for specific decisions on request.
    \end{itemize}
\end{itemize}

\subsection{ErisML integration}

ErisML runs as a library inside the MCP server. For each option:

\begin{enumerate}
  \item Parses EthicalFacts into internal structures.
  \item Calls the appropriate \textbf{domain Ethics Modules} (e.g., TriageEM) to compute moral vectors.
  \item Passes these vectors to the \textbf{governance engine} for veto and scalarization.
  \item Returns structured outputs plus explanation tags.
\end{enumerate}

Configuration allows multiple \textbf{domain modules} and \textbf{policy modules} to be active in the same profile.

\section{DEME--ErisML Gazebo Plugin}

\subsection{Plugin type and lifecycle}

The plugin is typically implemented as a \textbf{Gazebo world plugin} (for multi-agent triage) or a \textbf{model plugin} attached to the triage robot/agent. Its lifecycle:

\begin{enumerate}
  \item \textbf{Initialization}
    \begin{itemize}
      \item Reads configuration from SDF/INI/YAML:
        \begin{itemize}
          \item \texttt{mcp\_endpoint}
          \item \texttt{profile\_id}
          \item mapping from Gazebo models/topics to patient identities and attributes.
        \end{itemize}
      \item Establishes a persistent connection to the DEME MCP server.
    \end{itemize}

  \item \textbf{Per-update callback}
    \begin{itemize}
      \item At each control tick (e.g., 10--50 Hz), the plugin:
        \begin{itemize}
          \item Reads the agent's proposed action(s) from ROS topics or Gazebo messages.
          \item Constructs EthicalFacts for each option.
          \item Sends an evaluation request and waits for a bounded response time.
          \item Applies the response: vetoes or selected action.
        \end{itemize}
    \end{itemize}

  \item \textbf{Shutdown}
    \begin{itemize}
      \item Closes connections, flushes local logs,
      \item Optionally writes a summary report of decisions taken during the episode.
    \end{itemize}
\end{enumerate}

\subsection{Configuration example}

In an SDF world file:

\begin{lstlisting}[language=XML]
<plugin name="deme_erisml_governance"
        filename="libdeme_erisml_gazebo.so">
  <mcp_endpoint>tcp://localhost:8088</mcp_endpoint>
  <profile_id>hospital_triage_v2</profile_id>
  <agent_topic>/triage_agent/command</agent_topic>
  <governed_topic>/triage_agent/governed_command</governed_topic>
  <max_eval_latency_ms>20</max_eval_latency_ms>
</plugin>
\end{lstlisting}

The triage agent publishes desired commands to \texttt{/triage\_agent/command}; the plugin publishes \textbf{ethically governed commands} to \texttt{/triage\_agent/governed\_command} after consulting ErisML.

\section{Medical Triage Emergency Scenario}

\subsection{Environment}

The Gazebo world emulates an \textbf{emergency department (ED)} with:

\begin{itemize}
  \item An ICU area with a small number of \textbf{critical care beds}.
  \item A triage desk where a \textbf{triage agent} chooses which patient to admit next.
  \item Multiple patient models (A, B, C, \ldots) with:
    \begin{itemize}
      \item Vital signs (HR, BP, O\textsubscript{2}), lab results, SOFA scores.
      \item Demographic and social markers (age, housing status).
      \item Arrival times and event history.
    \end{itemize}
\end{itemize}

The triage agent's job is to \textbf{assign scarce ICU capacity} and schedule interventions.

\subsection{EthicalFacts extraction}

The plugin (or a helper node) transforms simulation state into EthicalFacts. For each candidate ``assign patient X to ICU'' action, we compute:

\begin{itemize}
  \item \texttt{physical\_harm\_if\_delayed} -- estimated mortality risk if patient waits.
  \item \texttt{expected\_benefit} -- probability of significant improvement if treated now.
  \item \texttt{disadvantaged\_status} -- boolean flag for structurally disadvantaged groups (e.g., unhoused, underinsured).
  \item \texttt{vip\_flag} -- boolean; should \emph{never} justify preferential treatment.
  \item \texttt{arrival\_time} -- for fairness and queueing rules.
  \item \texttt{epistemic\_confidence} -- reliability of risk estimates (e.g., presence of complete diagnostics).
  \item Legal and policy context flags (e.g., EMTALA applicability).
\end{itemize}

Example (pseudo-JSON):

\begin{lstlisting}
{
  "patient_id": "A",
  "physical_harm_if_delayed": 0.7,
  "expected_benefit": 0.85,
  "disadvantaged_status": true,
  "vip_flag": false,
  "arrival_time": 1700000000.12,
  "epistemic_confidence": 0.72
}
\end{lstlisting}

\subsection{Governance profile for triage}

A sample \texttt{hospital\_triage\_v2} profile might specify:

\begin{itemize}
  \item \textbf{Veto rules}
    \begin{itemize}
      \item Forbid any option where a VIP flag influences priority if clinical need is similar.
      \item Forbid actions that cause extreme harm when alternatives with substantially lower harm exist.
      \item Forbid discrimination across protected attributes.
    \end{itemize}

  \item \textbf{Lexicographic priorities}
    \begin{enumerate}
      \item Minimize expected harm (beneficence, non-maleficence).
      \item Among similarly effective options, maximize fairness to the worst-off.
      \item Maintain procedural legitimacy (follow institutional policies).
    \end{enumerate}
\end{itemize}

The DEME MCP server uses this profile to interpret the moral vectors produced by TriageEM.

\subsection{Example decision step}

In a particular simulation tick:

\begin{itemize}
  \item Patient A and C have similar clinical status; A is unhoused; C is a VIP donor.
  \item One ICU bed is available.
  \item The agent proposes three options:
    \begin{itemize}
      \item \texttt{assign\_A\_ICU1}, \texttt{assign\_B\_ICU1}, \texttt{assign\_C\_ICU1}.
    \end{itemize}
\end{itemize}

ErisML computes moral vectors:

\begin{table}[h]
\centering
\begin{tabular}{@{}lcccc@{}}
\toprule
Option & Harm $\downarrow$ & Rights & Fairness & Legitimacy \\
\midrule
\texttt{assign\_A\_ICU1} & 0.20 & 1.0 & 0.90 & 0.92 \\
\texttt{assign\_B\_ICU1} & 0.30 & 1.0 & 0.60 & 0.95 \\
\texttt{assign\_C\_ICU1} & 0.22 & 0.25 & 0.15 & 0.40 \\
\bottomrule
\end{tabular}
\end{table}

The profile:

\begin{itemize}
  \item \textbf{Vetoes} \texttt{assign\_C\_ICU1} because rights and fairness fall below thresholds.
  \item Compares A vs B lexicographically: A wins on harm and fairness.
  \item Returns winner = \texttt{assign\_A\_ICU1} and a reason code like \texttt{"EMTALA\_VIOLATION"} for C.
\end{itemize}

The Gazebo plugin enforces this:

\begin{itemize}
  \item Override any request to assign C.
  \item Apply an action to move A's model to the ICU bed.
  \item Log the governed decision so researchers can inspect trajectories and fairness statistics.
\end{itemize}

\subsection{Evaluating learning agents}

If a reinforcement learning agent controls triage decisions:

\begin{itemize}
  \item During training, the plugin can \textbf{mask vetoed actions} so the agent never receives positive reward for them.
  \item During deployment, the agent's proposed action is always filtered through the DEME governance layer.
  \item Researchers can examine whether different reward functions still converge to ethically acceptable policies under the same profile.
\end{itemize}

\section{Safety and Testing Strategy}

To use the DEME--ErisML plugin responsibly, we recommend:

\begin{enumerate}
  \item \textbf{Unit and property tests}
    \begin{itemize}
      \item Verify that the plugin correctly maps Gazebo state $\to$ EthicalFacts.
      \item Ensure vetoed options are never executed, even under communication delays.
    \end{itemize}

  \item \textbf{Scenario regression suites}
    \begin{itemize}
      \item Curate canonical triage scenarios:
        \begin{itemize}
          \item VIP vs disadvantaged patient with similar needs.
          \item Many low-risk vs one high-risk patient.
          \item Cases with uncertain diagnostics.
        \end{itemize}
      \item Run them under multiple governance profiles to confirm behavior.
    \end{itemize}

  \item \textbf{Stress testing}
    \begin{itemize}
      \item Vary simulation timestep, network latency to MCP, and number of concurrent triage calls.
      \item Confirm that the plugin fails \emph{safe} (e.g., conservative defaults) if the MCP server is unreachable or times out.
    \end{itemize}

  \item \textbf{Fairness and rights analysis}
    \begin{itemize}
      \item Analyze decision logs by patient group, arrival time, and clinical condition.
      \item Ensure that discrimination is not reintroduced inadvertently by perception or scoring models upstream of DEME.
    \end{itemize}
\end{enumerate}

\section{Integration with Learning Systems}

The governance plugin is designed to be \textbf{agent-agnostic}:

\begin{itemize}
  \item Classical planners, heuristic policies, and RL agents all produce candidate actions in the same way;
  \item DEME acts as a \textbf{safety and governance envelope} around these systems.
\end{itemize}

Typical patterns:

\begin{itemize}
  \item \textbf{Constrained exploration} -- During training, the plugin masks actions that violate the profile, effectively training in a safe action space.
  \item \textbf{Reward shaping} -- The plugin can inject additional negative rewards for near-veto actions to encourage policies that stay comfortably within acceptable regions.
  \item \textbf{Curriculum learning} -- Profiles can be tightened over time (e.g., from utilitarian to strongly fairness-weighted) while observing how policies adapt.
\end{itemize}

\section{Deployment Roadmap}

Although the current focus is simulation in Gazebo, the architecture is deliberately close to real robotic stacks:

\begin{enumerate}
  \item Gazebo + ROS2 simulation $\to$
  \item Hardware-in-the-loop testing (real sensors, simulated environment) $\to$
  \item Physical pilot deployments in non-clinical environments (e.g., triage training centers) $\to$
  \item Controlled clinical pilots once regulatory and ethical approvals are obtained.
\end{enumerate}

Because DEME MCP and ErisML are \textbf{network services}, the same governance engine can be reused across simulation and deployment: only the \textbf{EthicalFacts extraction} and \textbf{actuation interface} change.

\section{Conclusion}

The \textbf{DEME--ErisML Governance Plugin for Gazebo} provides a concrete path from abstract ethical theory to governed behavior in robotics simulations. By connecting Gazebo to a DEME MCP server running ErisML, we can:

\begin{itemize}
  \item Express hospital triage policies as explicit, versioned governance profiles,
  \item Enforce hard vetoes and principled rankings in real time,
  \item Study how learning agents behave under ethical constraints,
  \item Generate rich, auditable logs for ethics research, regulation and safety engineering.
\end{itemize}

The \textbf{Medical Triage Emergency} scenario demonstrates how such a system can operationalize fairness, rights and harm minimization in a domain where these values are non-negotiable. Future work includes publishing reference implementations of the plugin, expanding the library of triage scenarios, and integrating with broader open-source safety and ethics tooling for robotics.

\begin{thebibliography}{9}

\bibitem{bond2026ge}
A.~H.~Bond, \emph{Geometric Ethics: The Mathematical Structure of Moral Reasoning}, v1.14, Department of Computer Engineering, San Jos\'e State University, Spring 2026.

\bibitem{bond2026erisml}
A.~H.~Bond, ``ErisML/DEME v3.0: A Library for Governed, Foundation-Model-Enabled Agents with Tensorial Ethics,'' \url{https://github.com/ahb-sjsu/erisml-lib}, 2026.

\end{thebibliography}

\end{document}
