\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}

\begin{document}

\begin{center}
{\LARGE\bfseries Title Page}
\end{center}

\bigskip

\noindent\textbf{Title:} Geometric Economics: Heuristic Bounded Rationality and the Bond Geodesic Equilibrium

\bigskip

\noindent\textbf{Author:}\\
Andrew H. Bond\\
Department of Computer Engineering\\
San Jos\'{e} State University\\
San Jos\'{e}, CA 95192, USA\\
Email: \href{mailto:andrew.bond@sjsu.edu}{andrew.bond@sjsu.edu}\\
ORCID: \href{https://orcid.org/0009-0003-2599-6158}{0009-0003-2599-6158}

\bigskip

\noindent\textbf{Acknowledgments:}\\
The author thanks Lucas Thiele (UCLA) for conducting the independent replication study whose results are cited in this work.

\bigskip

\section*{Statements and Declarations}

\subsection*{Competing Interests}
The author has no relevant financial or non-financial interests to disclose.

\subsection*{Funding}
No funding was received for conducting this study.

\subsection*{Data Availability}
The reference implementation, empirical datasets, and all experimental scripts from the parent framework are publicly available at \url{https://github.com/ahb-sjsu/erisml-lib} and via the \texttt{erisml} package on PyPI.  The cross-lingual corpus data and scoring pipelines are included in the repository.  The economic calibration pipeline described in the paper (estimation of $\Sigma$ and $\beta_k$ from experimental economic data) has not yet been executed; the behavioral datasets required (e.g., ultimatum game, dictator game, public goods game data from standard experimental economics repositories) are publicly available.

\subsection*{Code Availability}
The parent framework's library is open-source under the MIT license at \url{https://github.com/ahb-sjsu/erisml-lib}.  The economic extension module described in the paper is planned but not yet implemented.

\subsection*{Author Contributions}
A.H.B. conceived the framework, developed the mathematical theory, designed the empirical calibration pipeline, and wrote the manuscript.

\subsection*{Use of AI Tools}
AI-assisted copy editing tools were used to improve readability and style during manuscript preparation.  All theoretical content, mathematical proofs, framework design, and scientific conclusions are the sole responsibility of the author.

\end{document}
