\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}

\begin{document}

\begin{center}
{\LARGE\bfseries Title Page}
\end{center}

\bigskip

\noindent\textbf{Title:} From Pathfinding to Field Theory: The Transition from $A^*$ on the Moral Manifold to the Moral Field Equation

\bigskip

\noindent\textbf{Author:}\\
Andrew H. Bond\\
Department of Computer Engineering\\
San Jos\'{e} State University\\
San Jos\'{e}, CA 95192, USA\\
Email: \href{mailto:andrew.bond@sjsu.edu}{andrew.bond@sjsu.edu}\\
ORCID: \href{https://orcid.org/0009-0003-2599-6158}{0009-0003-2599-6158}

\bigskip

\noindent\textbf{Acknowledgments:}\\
The author thanks Lucas Thiele (UCLA) for the independent replication study that strengthened the empirical basis of the Geometric Ethics framework. The transition from $A^*$ to field theory was catalyzed by a conversation with a colleague whose framing of the four-upgrade structure clarified the architecture of the research program. The author also thanks the anonymous reviewers for their constructive feedback.

\bigskip

\section*{Statements and Declarations}

\subsection*{Competing Interests}
The author has no relevant financial or non-financial interests to disclose.

\subsection*{Funding}
No funding was received for conducting this study.

\subsection*{Data Availability}
The reference implementation (ErisML v3.0), empirical datasets, and all experimental scripts are publicly available at \url{https://github.com/ahb-sjsu/erisml-lib} and via \texttt{pip install erisml-lib}. The CHSH Bell test data and cross-lingual passage datasets are included in the repository.

\subsection*{Code Availability}
The ErisML library (v3.0) is open-source under the MIT license, available on PyPI and GitHub (see above).

\subsection*{Author Contributions}
A.H.B. conceived the framework, developed the mathematical theory, conducted the CHSH falsification experiments, implemented the reference software, and wrote the manuscript.

\subsection*{Ethics Approval}
Not applicable. This study did not involve human participants, animal subjects, or identifiable personal data.

\subsection*{Consent to Participate}
Not applicable.

\subsection*{Consent for Publication}
Not applicable.

\subsection*{Use of AI Tools}
AI-assisted tools were used for copy editing and manuscript preparation. All theoretical content, mathematical proofs, empirical analyses, and scientific conclusions are the sole responsibility of the author.

\end{document}
