\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}

\begin{document}

\begin{center}
{\LARGE\bfseries Title Page}
\end{center}

\bigskip

\noindent\textbf{Title:} Geometric Ethics: Invariance, Conservation, and the Structure of Moral Evaluation

\bigskip

\noindent\textbf{Author:}\\
Andrew H. Bond\\
Department of Computer Engineering\\
San Jos\'{e} State University\\
San Jos\'{e}, CA 95192, USA\\
Email: \href{mailto:andrew.bond@sjsu.edu}{andrew.bond@sjsu.edu}\\
ORCID: \href{https://orcid.org/0009-0003-2599-6158}{0009-0003-2599-6158}

\bigskip

\noindent\textbf{Acknowledgments:}\\
The author thanks Lucas Thiele (UCLA) for conducting the independent replication study that strengthened the empirical basis of this work. The author also thanks the anonymous reviewers for their constructive feedback.

\bigskip

\section*{Statements and Declarations}

\subsection*{Competing Interests}
The author has no relevant financial or non-financial interests to disclose.

\subsection*{Funding}
No funding was received for conducting this study.

\subsection*{Data Availability}
The reference implementation (ErisML v3.0), empirical datasets, and all experimental scripts are publicly available at \url{https://github.com/ahb-sjsu/erisml-lib} and via \texttt{pip install erisml-lib}. The Dear Abby corpus analysis data and cross-lingual passage datasets are included in the repository.

\subsection*{Code Availability}
The ErisML library (v3.0) is open-source under the MIT license, available on PyPI and GitHub (see above).

\subsection*{Author Contributions}
A.H.B. conceived the framework, developed the mathematical theory, conducted all empirical analyses, implemented the reference software, and wrote the manuscript.

\subsection*{Use of AI Tools}
AI-assisted copy editing tools were used to improve readability and style during manuscript preparation. All theoretical content, mathematical proofs, empirical analyses, and scientific conclusions are the sole responsibility of the author.

\end{document}
